Understanding the Landscape of Childlessness in the U.S.

### 2024-03-18 outputs/paper/paper.qmd
Context arragment change

# Introduction {#sec-intro}
The American family landscape is undergoing a significant transformation, marked by a growing trend of childlessness. Traditionally viewed as a societal norm, having children now becomes a choice [@voluntaryChild]. We examine factors such as age, gender, education, socioeconomic class, health conditions and marital status, and how they affect childbearing rates over the past 50 years.

@@

Our estimand is the relationship between age, gender, socio-economic class, educational attainment, health status, marital status and the rate of childlessness in the United States from 1972 to 2022. Through our analysis, we found that childless rate is consistently on the rise from 1972 to 2022, with correlation coefficient of 0.73 between childless percentage and time (or a 6.99% increase from 1972 to 2022). Factors that directly contribute to this rise are higher educational attainment, higher socio-economic class,lower gender disparity, better health status and dissolution of marriages through divorces, separation and being widowed.

@@

Each variable used—age, class, educational attainment, health, marital status, and gender—is  quantified through direct questioning to accurately capture respondents' backgrounds. For instance, the General Social Survey (GSS) obtains information on marital status through questions such as, "Are you currently married, widowed, divorced, separated, or have you never been married?", with responses categorized into predefined groups for further analysis [@gssCodebook]. On ther other hand, instead of relying on income sources, socio-economic class is subjectively assessed based on the respondents' answers to this question, "If you were asked to use one of four names for your social class, which would you say you belong in: the lower class, the working class, the middle class, or the upper class?". Some are continuous variable, like age. With respondents' various age groups documented, we can understand the generational disparities in childbearing patterns. Systematic and data-driven, these measurement methods ensures that all data faithfully reflects the American population's childbearing tendencies.

@@

However, this multimodal survey approach also presents several challenges in collecting respondents' opinions on topics such as childbirth and childbearing tendencies, given the private nature of these subjects. With the easy formatting and non-required nature of these surveys, biases like acquiescence bias (only positive responses) might occur because respondents aren't confronted in a face-to-face settings, where many will just opt a neutral response [@webSurvey]. Therefore, we carefully select various variables to to ensure the authenticity of the public's background on the topic of childlessness, avoiding having the data being too influenced by the changes in surveying methods.

@@

Thus, there is a consistent rise in childlessness across all age groups in the United States from 1972 to 2022. Notably, the most significant rise is observed among younger adults (18-34 and 35-49). It is important to acknowledge, however, that the data series has gaps for several years within this timeframe (1979, 1981, 1992, 1995, 1997, 1999, 2001, 2003, 2005, 2007, 2009, 2011, 2013, 2015, 2017, and 2019). These missing data points can potentially limit the generalization of the observed trends and hinder a comprehensive understanding of childlessness over time.

@@

The data reveals a gradual increase in childlessness for both genders over the past five decades. However, females consistently exhibit a lower percentage of childlessness compared to males throughout the entire period. While the gap appears to be slightly narrowing over time, females still have a lower percentage by around 8 percentage points in 2022 (approximately 28.32% for females and 36.09% for males).

@@

@@

@@

### Overview of Educational Attainment's Impact on Childlessness

@@

In addition, additional data sets from NORC [@gss] were also leveraged to understand the context of childlessness in America from 1972-2022. @tbl-nochild-health and @tbl-nochild-marital present an interesting connection between childlessness and health conditions and marital status.

* Health Conditions: This variable, measured by medical check-ups of individuals who have been surveyed, will be used to clarify the context behind the childlessness of those with varying health statuses.
* Marital Status: This variable enables us to see if marriage is a determining factor in childbearing decisions.

@@

\newpage
### Overview of Health Status's Impact on Childlessness

@tbl-nochild-marital explores people's marital status and childlessness. @fig-nochild-health-marital then illustrates this data, exploring the marital status of people who are childless over a span of 50 years in the United States (1972-2022). The graph contains trend lines of varying marital statuses-Divorced, Married, Never married, Separated and Widowed-and how each has changed over the years in terms of childbearing status.

@@

Survey data may be susceptible to measurement errors due to various factors like social desirability bias, memory lapses, or misinterpretations of questions. While the GSS implements measures to minimize these errors, they cannot be entirely eliminated.

@@

Overall, the data shows an increase in childlessness across all health levels and conditions approaching 2020. As mentioned, @fig-nochild-health-marital shows that people with excellent health have the highest percentage of childlessness, having a 10.49% increase throughout the 50-year period (28.26% in 1972 to 38.75% in 2022). People with good and fair health exhibit a similar rate of childlessness in 2022, at about 30%. However, people with poor health, once having the lowest rate of childlessness (at 12.27% in 1972), now increases to 20.40% in 2022, which translates to a substantial 8.13% increase. Overall, on average, across all health levels from 1972-2022, @fig-avg-one shows that percentage of childlessness increases at about 6.836%.

@@

As for marital status, @fig-nochild-health-marital demonstrates that a significant portion of people who have never married are childless. However, more and more people are reluctant to get married but still have children. This is observable in the childlessness percentage decrease of people who never marry, decreasing from 96.73% in 1972 to 72.53% in 2022. For people who are divorced, separated and widowed, this percentage is on the rise fluctuates, displaying the complex dynamics between these relationship statuses and childbearing. For married people though, this childlessness percentage also decreases, from 10.51% in 1972 to 13.97% in 2022. Overall, on average, across all marital status from 1972-2022, @fig-avg-one shows that percentage of childlessness decreases at about 7.39%.

@@

## Between Socio-economic Class, Educational Attainment and Childlessness

While all classes show an increase of childlessness in @fig-nochild-class-edu, the data suggests the most significant rise occurred among the lower class, with their percentage of childless individuals nearly doubling from approximately 17.16% in 1972 to around 34.46% in 2022, marking a substantial increase of approximately 101.04% over the specified period. The working class also experienced a significant increase, from around 21.83% in 1972 to approximately 31.44% in 2022, indicating a rise of roughly 43.98%. Additionally, the middle class saw a slight increase, from approximately 28.24% in 1972 to about 30.09% in 2022, representing a rise of approximately 6.55%. Similarly, the upper class saw an increase, with their percentage of childless individuals climbing from approximately 20.24% in 1972 to about 27.87% in 2022, indicating an increase of roughly 37.71%.

@@

Similarly, data reveals a distinct pattern across educational groups. Individuals with a high school diploma consistently exhibit the highest percentage of childlessness throughout the entire period, ranging from around 28.15% in 1972 to 34.56% in 2022. Conversely, the percentage of childless individuals with a college and higher degree has decreased from 32.85% in 1972 to 29.84% in 2022. The trend for individuals with a with less than a high school education increased from 16.97% in 1972 to 22.90% in 2022.

@@

@@

\newpage
@fig-overall graphs a trend-line of overall childlessness rate throughout the years. With a correlation coefficient between "Year" and "Percentage" of 0.73, it's highly likely that as time progresses, childlessness is on the rise. This is evident in the 6.99% increase in childlessness from 1972 to 2022.

To account for the general increase of childlessness from 1972 to 2022, there are several crucial events. First, in @fig-overall, the first surge of childlessness occurs from 1986 onwards till 1900 (27.71% to 30.26%, a 2.55% increase). This was when the AIDs pandemic occurred, where there was much fear and health concerns of HIV transmission in having sexual intercourse. The second surge of childlessness is observable from 2000 onwards till 2006 (27.76% to 28.94%, a 1.18% increase). This was the time period marked by fear of terrorism when 9/11 happened, where heightened feelings of vulnerability may have influenced people's decisions of expanding their family. The third surge of childlessness from 2008 till 2014 is drastic due to the occurrence of the Great Recession (27.25% to 30.75%, a 3% increase). Financial instability then is a contributing factor in determining childlessness. The last surge occured from 2018 to 2021 when the COVID-19 was in place (29.69% to 31.67%, a 1.98% increase).

>>>>

# Introduction {#sec-intro}
The American family landscape is undergoing a significant transformation, marked by a growing trend of childlessness. Traditionally viewed as a societal norm, having children now becomes a choice [@voluntaryChild]. We examine factors such as age, gender, education, socioeconomic class, health conditions and marital status, and how they have affected childbearing rates over the past 50 years.

@@

Our estimand is the relationship between age, gender, socioeconomic class, educational attainment, health status, marital status and the rate of childlessness in the United States from 1972 to 2022. Through our analysis, we found that the childless rate is consistently on the rise from 1972 to 2022, with a correlation coefficient of 0.73 between the childless percentage and time (or a 6.99% increase from 1972 to 2022). Factors that directly contribute to this rise are higher educational attainment, higher socioeconomic class, lower gender disparity, better health status and dissolution of marriages through divorces, separation and being widowed.

@@

Each variable used—age, class, educational attainment, health, marital status, and gender—is quantified through direct questioning to accurately capture respondents' backgrounds. For instance, the General Social Survey (GSS) obtains information on marital status through questions such as, "Are you currently married, widowed, divorced, separated, or have you never been married?", with responses categorized into predefined groups for further analysis [@gssCodebook]. On the other hand, instead of relying on income sources, socioeconomic class is subjectively assessed based on the respondents' answers to this question, "If you were asked to use one of four names for your social class, which would you say you belong in: the lower class, the working class, the middle class, or the upper class?". Some are continuous variables, like age. With respondents' various age groups documented, we can understand the generational disparities in childbearing patterns. Systematic and data-driven, these measurement methods ensure that all data faithfully reflects the American population's childbearing tendencies.

@@

However, this multimodal survey approach also presents several challenges in collecting respondents' opinions on topics such as childbirth and childbearing tendencies, given the private nature of these subjects. With the easy formatting and non-required nature of these surveys, biases like acquiescence bias (only positive responses) might occur because respondents aren't confronted in a face-to-face setting, where many will just opt for a neutral response [@webSurvey]. Therefore, we carefully select various variables to to ensure the authenticity of the public's background on the topic of childlessness, avoiding having the data being too influenced by the changes in surveying methods.

@@

Thus, there is a consistent rise in childlessness across all age groups in the United States from 1972 to 2022. The most significant rise is observed among younger adults (18-34 and 35-49). However, it is important to acknowledge that the data series has gaps for several years within this timeframe (1979, 1981, 1992, 1995, 1997, 1999, 2001, 2003, 2005, 2007, 2009, 2011, 2013, 2015, 2017, and 2019). These missing data points can potentially limit the generalization of the observed trends and hinder a comprehensive understanding of childlessness over time.

@@

The data reveals a gradual increase in childlessness for both genders over the past five decades. However, females consistently exhibit a lower percentage of childlessness compared to males throughout the entire period. While the gap appears to be slightly narrowing over time, females still have a lower percentage by around eight percentage points in 2022 (approximately 28.32% for females and 36.09% for males).

@@

\newpage
### Overview of Educational Attainment's Impact on Childlessness

@@

In addition, additional data sets from NORC [@gss] were also leveraged to understand the context of childlessness in America from 1972-2022. @tbl-nochild-health and @tbl-nochild-marital present an interesting connection between childlessness, health conditions and marital status.

* **Health Conditions:** This variable, measured by medical check-ups of individuals who have been surveyed, will be used to clarify the context behind the childlessness of those with varying health statuses.
* **Marital Status:** This variable enables us to see if marriage is a determining factor in childbearing decisions.

### Overview of Health Status's Impact on Childlessness

@@

@tbl-nochild-marital explores people's marital status and childlessness. @fig-nochild-health-marital then illustrates this data, exploring the marital status of people who are childless over a span of 50 years in the United States (1972-2022). The graph contains trend lines of varying marital statuses-divorced, married, never married, separated and widowed-and how each has changed over the years in terms of childbearing status.

@@

@@

@@

Survey data may be susceptible to measurement errors due to various factors, such as social desirability bias, memory lapses, or misinterpretations of questions. While the GSS implements measures to minimize these errors, they cannot be entirely eliminated.

@@

Overall, the data shows an increase in childlessness across all health levels and conditions approaching 2020. As mentioned, @fig-nochild-health-marital shows that people with excellent health have the highest percentage of childlessness, having a 10.49% increase throughout the 50-year period (28.26% in 1972 to 38.75% in 2022). People with good and fair health exhibit a similar rate of childlessness in 2022, at about 30%. However, people with poor health, once having the lowest rate of childlessness (at 12.27% in 1972), now increases to 20.40% in 2022, which translates to a substantial 8.13% increase. Overall, on average, across all health levels from 1972-2022, @fig-avg-one shows that the percentage of childlessness increases at about 6.836%.

@@

As for marital status, @fig-nochild-health-marital demonstrates that a significant portion of people who have never married are childless. However, more and more people are reluctant to get married but still have children. This is observable in the childlessness percentage decrease of people who never marry, decreasing from 96.73% in 1972 to 72.53% in 2022. For people who are divorced, separated and widowed, this percentage is on the rise fluctuates, displaying the complex dynamics between these relationship statuses and childbearing. For married people though, this childlessness percentage also decreases, from 10.51% in 1972 to 13.97% in 2022. Overall, on average, across all marital statuses from 1972-2022, @fig-avg-one shows that the percentage of childlessness decreases to about 7.39%.

@@

## Between Socioeconomic Class, Educational Attainment and Childlessness

While all classes show an increase in childlessness in @fig-nochild-class-edu, the data suggests the most significant rise occurred among the lower class, with their percentage of childless individuals nearly doubling from approximately 17.16% in 1972 to around 34.46% in 2022, marking a substantial increase of approximately 101.04% over the specified period. The working class also experienced a significant increase, from around 21.83% in 1972 to approximately 31.44% in 2022, indicating a rise of roughly 43.98%. Additionally, the middle class saw a slight increase, from approximately 28.24% in 1972 to about 30.09% in 2022, representing a rise of approximately 6.55%. Similarly, the upper class saw an increase, with their percentage of childless individuals climbing from approximately 20.24% in 1972 to about 27.87% in 2022, indicating an increase of roughly 37.71%.

@@

Similarly, data reveals a distinct pattern across educational groups. Individuals with a high school diploma consistently exhibit the highest percentage of childlessness throughout the entire period, ranging from around 28.15% in 1972 to 34.56% in 2022. Conversely, the percentage of childless individuals with a college and higher degree has decreased from 32.85% in 1972 to 29.84% in 2022. The trend for individuals with less than a high school education increased from 16.97% in 1972 to 22.90% in 2022.

@@

@fig-overall graphs a trend line of the overall childlessness rate throughout the years. With a correlation coefficient between "Year" and "Percentage" of 0.73, it's highly likely that as time progresses, childlessness is on the rise. This is evident in the 6.99% increase in childlessness from 1972 to 2022.

To account for the general increase in childlessness from 1972 to 2022, there are several crucial events. First, in @fig-overall, the first surge of childlessness occurs from 1986 onwards till 1900 (27.71% to 30.26%, a 2.55% increase). This was when the AIDs pandemic occurred, when there was much fear and health concerns about HIV transmission in having sexual intercourse. The second surge of childlessness is observable from 2000 onwards till 2006 (27.76% to 28.94%, a 1.18% increase). This was the time period marked by fear of terrorism when 9/11 happened, where heightened feelings of vulnerability may have influenced people's decisions of expanding their family. The third surge of childlessness from 2008 till 2014 is drastic due to the occurrence of the Great Recession (27.25% to 30.75%, a 3% increase). Financial instability then is a contributing factor in determining childlessness. The last surge occurred from 2018 to 2021 when COVID-19 was in place (29.69% to 31.67%, a 1.98% increase).

@@
